CF Registry Analysis

### 2025-09-16 Christine Chan/CF Registry/cf_registry_CFRD_prediction.qmd
updated plot for Christine

```{r setup}
#| include: false
library(survival)
library(nlme)
library(JM)
library(dplyr)
library(tidyr)
library(arsenal)
library(knitr)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

demo <- read.csv("./Data_Raw/CFF19_DemogCFDiag_Del1.csv",

@@

  "White", "Black or African American", "American Indian or Alaska Native",
  "Asian", "Native Hawaiian or Other Pacific Islander", "Some other race"

@@

demo$Race <- factor(demo$Race,

@@

    "White", "Black or African American", "American Indian or Alaska Native",
    "Asian", "More than one race", "Native Hawaiian or Other Pacific Islander",

@@

    "White", "Black or African American", "Other race", "Other race",
    "More than one race", "Other race", "Other race"

@@

demo$Hispanicrace <- factor(demo$Hispanicrace,

@@

demo$F508 <- factor(demo$F508,

@@

annual$A_VX809comb <- factor(annual$A_VX809comb,
  levels = 0:1, labels = c("No", "Yes")

@@

  breaks = c(0, 1, 2, 3, Inf), right = F

@@

annual$A_IsOnEnzymes <- factor(annual$A_IsOnEnzymes,
  levels = 1:0, labels = c("Yes", "No")

@@

  Hispanicrace = "Hispanic", A_Vx770 = "Use of CFTR modulator Vx770",

@@

  A_FEV1_pct_predicted = "FEV1 percent predicted", MutClass = "Mutation class",

@@

  A_IsOnEnzymes = "Pancreatic enzyme use", bmi = "BMI value"

@@

  "Gender", "Race", "Hispanicrace", "A_IsOnEnzymes",
  "NumPulmExacerbation", "A_FEV1_pct_predicted", "bmi"

@@

    eDWID, ReviewYear, Age_YrEnd, A_hgba1c, all_of(predictors), cfrd_registry

@@

```{r,results='asis'}
t <- registry[registry$eDWID %in% reg_include, ] %>%
  group_by(eDWID) %>%
  summarise(n())
t2 <- registry[registry$eDWID %in% reg_include, ] %>%
  group_by(eDWID) %>%
  filter(row_number() == 1)
t1 <- tableby(cfrd_registry ~ Age_YrEnd + Gender + Race + Hispanicrace + A_IsOnEnzymes + NumPulmExacerbation + A_FEV1_pct_predicted + bmi, data = t2)
summary(t1)
```

@@

```{r}
# Formulas
cox_f <- as.formula(paste0(
  "Surv(time = Age_YrEnd, event = cfrd_registry)~",
  paste0(predictors, collapse = "+")
))
lme_f <- as.formula(paste0(
  "A_hgba1c~Age_YrEnd+",
  paste0(predictors, collapse = "+")
))
# Fit Cox model first, and then pull the clusters
cox_fit <- coxph(cox_f,
  cluster = eDWID, x = T, model = T,
  data = surv_reg[surv_reg$eDWID %in% reg_include, ]
)
lme_fit_a1c <- lme(lme_f,
  random = ~ Age_YrEnd | eDWID, na.action = na.omit,
  data = registry[registry$eDWID %in% reg_include, ]
)
# Combine
joint_fit_value <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd")
# joint_fit_slope <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd",parameterization ="slope")
# joint_fit_both <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd",parameterization ="both")
s <- summary(joint_fit_value)
long <- data.frame(s$`CoefTable-Long`)
long$p.value <- format.pval(long$p.value, eps = 0.001, digits = 3)
event <- data.frame(s$`CoefTable-Event`)
event$Value <- exp(event$Value)
rownames(event)[nrow(event)] <- "shape"
event$p.value <- format.pval(event$p.value, eps = 0.001, digits = 3)
# summary(joint_fit_slope)
# summary(joint_fit_both)
kable(long, caption = "Longitudinal Process", digits = 3)
kable(event, caption = "Event Process", digits = 3)
```

>>>>

```{r setup}
#| include: false
library(survival)
library(nlme)
library(JM)
library(dplyr)
library(tidyr)
library(arsenal)
library(knitr)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

demo <- read.csv(
  "./Data_Raw/CFF19_DemogCFDiag_Del1.csv",

@@

  "White",
  "Black or African American",
  "American Indian or Alaska Native",
  "Asian",
  "Native Hawaiian or Other Pacific Islander",

@@

demo$Race <- factor(
  demo$Race,

@@

    "White",
    "Black or African American",
    "American Indian or Alaska Native",
    "Asian",
    "More than one race",
    "Native Hawaiian or Other Pacific Islander",

@@

    "White",
    "Black or African American",
    "Other race",
    "Other race",
    "More than one race",
    "Other race",
    "Other race"

@@

demo$Hispanicrace <- factor(
  demo$Hispanicrace,

@@

demo$F508 <- factor(
  demo$F508,

@@

annual$A_VX809comb <- factor(
  annual$A_VX809comb,
  levels = 0:1,
  labels = c("No", "Yes")

@@

  breaks = c(0, 1, 2, 3, Inf),
  right = F

@@

annual$A_IsOnEnzymes <- factor(
  annual$A_IsOnEnzymes,
  levels = 1:0,

@@

  Hispanicrace = "Hispanic",
  A_Vx770 = "Use of CFTR modulator Vx770",

@@

  A_FEV1_pct_predicted = "FEV1 percent predicted",
  MutClass = "Mutation class",

@@

  A_IsOnEnzymes = "Pancreatic enzyme use",
  bmi = "BMI value"

@@

  "Gender",
  "Race",
  "Hispanicrace",
  "A_IsOnEnzymes",
  "NumPulmExacerbation",
  "A_FEV1_pct_predicted",
  "bmi"

@@

    eDWID,
    ReviewYear,
    Age_YrEnd,
    A_hgba1c,
    all_of(predictors),
    cfrd_registry

@@

```{r,results='asis'}
t <- registry[registry$eDWID %in% reg_include, ] %>%
  group_by(eDWID) %>%
  summarise(n())
t2 <- registry[registry$eDWID %in% reg_include, ] %>%
  group_by(eDWID) %>%
  filter(row_number() == 1)
t1 <- tableby(
  cfrd_registry ~
    Age_YrEnd +
      Gender +
      Race +
      Hispanicrace +
      A_IsOnEnzymes +
      NumPulmExacerbation +
      A_FEV1_pct_predicted +
      bmi,
  data = t2
)
summary(t1)
```

@@

```{r}
# Formulas
cox_f <- as.formula(paste0(
  "Surv(time = Age_YrEnd, event = cfrd_registry)~",
  paste0(predictors, collapse = "+")
))
lme_f <- as.formula(paste0(
  "A_hgba1c~Age_YrEnd+",
  paste0(predictors, collapse = "+")
))
# Fit Cox model first, and then pull the clusters
cox_fit <- coxph(
  cox_f,
  cluster = eDWID,
  x = T,
  model = T,
  data = surv_reg[surv_reg$eDWID %in% reg_include, ]
)
lme_fit_a1c <- lme(
  lme_f,
  random = ~ Age_YrEnd | eDWID,
  na.action = na.omit,
  data = registry[registry$eDWID %in% reg_include, ]
)
# Combine
joint_fit_value <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd")
# joint_fit_slope <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd",parameterization ="slope")
# joint_fit_both <- jointModel(lme_fit_a1c, cox_fit, timeVar = "Age_YrEnd",parameterization ="both")
s <- summary(joint_fit_value)
long <- data.frame(s$`CoefTable-Long`)
long$p.value <- format.pval(long$p.value, eps = 0.001, digits = 3)
event <- data.frame(s$`CoefTable-Event`)
event$Value <- exp(event$Value)
rownames(event)[nrow(event)] <- "shape"
event$p.value <- format.pval(event$p.value, eps = 0.001, digits = 3)
# summary(joint_fit_slope)
# summary(joint_fit_both)
kable(long, caption = "Longitudinal Process", digits = 3)
kable(event, caption = "Event Process", digits = 3)
```

```{r}
nd <- joint_fit_value$data
nd$pred <- predict(
  joint_fit_value,
  registry[registry$eDWID %in% reg_include, ],
  type = "Subject",
  idVar = "eDWID"
)
```

### 2025-09-16 Christine Chan/CF Registry/cf_registry_survival_analysis.qmd
updated plot for Christine

```{r}
# Per reviewer, out of X individuals with a diagnosis of CFRD, X number had an
# OGTT (and X had an A1c, but no OGTT) that year or the year prior.
# This is super slow but idc idc
cfrd_only <- apply(first_cfrd, 1, function(r) {
    id <- as.numeric(r["eDWID"])
    year <- as.numeric(r["first_cfrd"])
    d <- annual[annual$eDWID == id & annual$ReviewYear %in% c(year, year - 1), ]
    a1c <- as.numeric(last(d["A_hgba1c"], na_rm = TRUE))
    ogtt0hr <- as.numeric(last(d["A_ogttfast_bloodglucose"], na_rm = TRUE))
    ogtt2hr <- as.numeric(last(d["A_twohour_bloodglucose"], na_rm = TRUE))
    return(c(id, year, a1c, ogtt0hr, ogtt2hr))
})
cfrd_only <- data.frame(t(cfrd_only))
colnames(cfrd_only) <- c("id", "year", "a1c", "ogtt0hr", "ogtt2hr")
cfrd_only <- left_join(
    cfrd_only,
    annual %>% mutate(eDWID = as.numeric(eDWID)),
    by = join_by(id == eDWID, year == ReviewYear)
)

cfrd_only$a1c_ogtt <- apply(cfrd_only, 1, function(r) {
    not_missing <- paste(
        which(!is.na(r[c("a1c", "ogtt0hr", "ogtt2hr")])),
        collapse = "_"
    )
    return(not_missing)
})
cfrd_only$a1c_ogtt <- factor(
    cfrd_only$a1c_ogtt,
    levels = c("", "1", "1_2", "1_2_3", "1_3", "2", "2_3", "3"),
    labels = c(
        "Neither",
        "HbA1c Only",
        "OGTT",
        "OGTT",
        "OGTT",
        "OGTT",
        "OGTT",
        "OGTT"
    )
)
cfrd_only %>%
    select(
        a1c_ogtt,
        Age_YrEnd,
        A_weight,
        A_height,
        A_bmivalue,
        A_bmipercentile,
        A_FEV1_pct_predicted,
        A_FVC_pct_predicted,
        Race,
        Hispanicrace,
        MutClass,
        a1c,
        A_ogttfast_bloodglucose,
        A_twohour_bloodglucose,
        A_IsOnEnzymes,
        liver_disease,
        A_cultureresults3
    ) %>%
    tbl_summary(
        by = a1c_ogtt,
        missing_text = "Missing",
        statistic = list(all_continuous() ~ "{mean} ({sd})")
    ) %>%
    add_overall() %>%
    add_p(test = list(all_continuous() ~ "oneway.test")) %>%
    modify_caption("**All Groups**")
cfrd_only %>%
    filter(a1c_ogtt %in% c("OGTT", "HbA1c Only")) %>%
    mutate(a1c_ogtt = droplevels(a1c_ogtt)) %>%
    select(
        a1c_ogtt,
        Age_YrEnd,
        A_weight,
        A_height,
        A_bmivalue,
        A_bmipercentile,
        A_FEV1_pct_predicted,
        A_FVC_pct_predicted,
        Race,
        Hispanicrace,
        MutClass,
        a1c,
        A_ogttfast_bloodglucose,
        A_twohour_bloodglucose,
        A_IsOnEnzymes,
        liver_disease,
        A_cultureresults3
    ) %>%
    tbl_summary(
        by = a1c_ogtt,
        missing_text = "Missing",
        statistic = list(all_continuous() ~ "{mean} ({sd})")
    ) %>%
    add_overall() %>%
    add_p(test = list(all_continuous() ~ "oneway.test")) %>%
    modify_caption("**OGTT vs. HbA1c Only**")
```

>>>>

```{r}
# Per reviewer, out of X individuals with a diagnosis of CFRD, X number had an
# OGTT (and X had an A1c, but no OGTT) that year or the year prior.
# This is super slow but idc idc
cfrd_only <- apply(first_cfrd, 1, function(r) {
  id <- as.numeric(r["eDWID"])
  year <- as.numeric(r["first_cfrd"])
  d <- annual[annual$eDWID == id & annual$ReviewYear %in% c(year, year - 1), ]
  a1c <- as.numeric(last(d["A_hgba1c"], na_rm = TRUE))
  ogtt0hr <- as.numeric(last(d["A_ogttfast_bloodglucose"], na_rm = TRUE))
  ogtt2hr <- as.numeric(last(d["A_twohour_bloodglucose"], na_rm = TRUE))
  return(c(id, year, a1c, ogtt0hr, ogtt2hr))
})
cfrd_only <- data.frame(t(cfrd_only))
colnames(cfrd_only) <- c("id", "year", "a1c", "ogtt0hr", "ogtt2hr")
cfrd_only <- left_join(
  cfrd_only,
  annual %>% mutate(eDWID = as.numeric(eDWID)),
  by = join_by(id == eDWID, year == ReviewYear)
)
cfrd_only$a1c_ogtt <- apply(cfrd_only, 1, function(r) {
  not_missing <- paste(
    which(!is.na(r[c("a1c", "ogtt0hr", "ogtt2hr")])),
    collapse = "_"
  )
  return(not_missing)
})
cfrd_only$a1c_ogtt <- factor(
  cfrd_only$a1c_ogtt,
  levels = c("", "1", "1_2", "1_2_3", "1_3", "2", "2_3", "3"),
  labels = c(
    "Neither",
    "HbA1c Only",
    "OGTT",
    "OGTT",
    "OGTT",
    "OGTT",
    "OGTT",
    "OGTT"
  )
)
cfrd_only %>%
  select(
    a1c_ogtt,
    Age_YrEnd,
    A_weight,
    A_height,
    A_bmivalue,
    A_bmipercentile,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_cultureresults3
  ) %>%
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  add_p(test = list(all_continuous() ~ "oneway.test")) %>%
  modify_caption("**All Groups**")
cfrd_only %>%
  filter(a1c_ogtt %in% c("OGTT", "HbA1c Only")) %>%
  mutate(a1c_ogtt = droplevels(a1c_ogtt)) %>%
  select(
    a1c_ogtt,
    Age_YrEnd,
    A_weight,
    A_height,
    A_bmivalue,
    A_bmipercentile,
    A_FEV1_pct_predicted,
    A_FVC_pct_predicted,
    Race,
    Hispanicrace,
    MutClass,
    a1c,
    A_ogttfast_bloodglucose,
    A_twohour_bloodglucose,
    A_IsOnEnzymes,
    liver_disease,
    A_cultureresults3
  ) %>%
  tbl_summary(
    by = a1c_ogtt,
    missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  add_p(test = list(all_continuous() ~ "oneway.test")) %>%
  modify_caption("**OGTT vs. HbA1c Only**")
# Plot by diagnosis type
plot_df = cfrd_only %>%
  select(id, a1c_ogtt) %>%
  rename(eDWID = id)
plot_df = left_join(
  annual %>% mutate(eDWID = as.numeric(eDWID)),
  plot_df,
  by = join_by(eDWID)
)
plot_df = plot_df %>%
  group_by(eDWID) %>%
  mutate(`Years to CFRD` = ReviewYear - first_cfrd)
ggplot(
  plot_df %>% filter(!is.na(a1c_ogtt)),
  aes(
    y = A_FEV1_pct_predicted,
    x = `Years to CFRD`,
    color = a1c_ogtt,
    group = a1c_ogtt
  )
) +
  xlim(c(-10, 10)) +
  geom_point(alpha = 0.1, shape = ".") +
  geom_smooth(formula = y ~ s(x, bs = "cr")) +
  theme_bw() +
  theme(legend.title = element_blank())
```
